---
format: html
editor: visual
filters: 
  - webr
title: "Introduction to GGPlot"
---

```{webr-r}
#| context: setup
#| label: input data

library(tidyverse)
library("palmerpenguins")
library(naniar)

penguin<- tibble('species'=penguins$species, 'island'=penguins$island, 'bill_length' = penguins$bill_length_mm, 'bill_depth' = penguins$bill_depth_mm, 'flipper_length'=penguins$flipper_length_mm, 'body_mass'=penguins$body_mass_g, 'year'=penguins$year, 'sex'=penguins$sex)

knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(echo = TRUE)
```

## An Introduction to GGPlot

In this tutorial, we'll be looking at how to create plots using the package `ggplot`.

We'll start by learning about how `ggplot` works.

### Syntax

First, let's consider the syntax when we write code with ggplot. There are many commands that can be used within the `ggplot()` function. They are all connected using a `+`.

General syntax for `ggplot()`:

-   `ggplot(data, aes(x=variable1, y=variable2))` `+`

    `geom_x()`

    OR

-   `dataframe` `|>`

    `ggplot(aes(x=variable1, y=variable2))` `+`

    `geom_x()`

Specifying variables: X is the explanatory variable and Y is the outcome variable

### Aesthetics

Within ggplot code, you can specify the aesthetics with various variables including color, size, variables, number of bins (in a histogram), etc.

For example the following code specifies the x & y variables, point size & color:

`ggplot(dataframe, aes(x=month, y=temperature) + geom_point(size=5, color=red)`

### Geometry

You can then include geometry including:

-   `geom_line()`: Connects all the points in a plot

-   `geom_smooth(method="lm", se=FALSE)`: Creates a line of best fit

-   `geom_point()`: Creates points (scatterplot) for each data point

-   `geom_bar()` : Creates a bar graph, aggregates data for you

    -   Syntax: `geom_bar(position="___")` Can choose "stack" (bars on top of each other), "dodge" (bars side by side) or "fill" (bars stacked, scaled to 100%).

-   `geom_col()`: Creates a bar graph with *pre-aggregated* data

-   `geom_histogram()`: Creates a histogram

    -   Syntax: `geom_histogram(bins=X)` Specify number of bins

-   `geom_density()`: Creates a density graph

    -   Syntax: `geom_density(adjust=X)` X is a ratio, how smooth the graph is

-   `labs()` : Add a title and axes labels to your graph

-   And more!

## Other functions to use with GGplot

You can use the pipe to incorporate other tidyverse functions to organize your data before plotting.

## Our Dataset

We will work with a dataset with measurements for 3 different species of penguins on islands in the Palmer Archipelago, Antarctica. We'll be focusing on the measurements for species, location, bill length (mm), bill depth (mm), flipper length (mm), body mass (g), sex, and year the data was taken. Here's more [information](https://cran.r-project.org/web/packages/palmerpenguins/readme/README.html) about the [data set](https://cran.r-project.org/web/packages/palmerpenguins/readme/README.html)!

Here is a preview of the data you will be working with:

```{webr-r}
penguin
```

## Exercises

Here are some practice exercises. You can check if your code works using the "check your work" button after each problem. 

If you get stuck, you can also click for hints and for an example answer. Remember, there's many ways to solve each problem, so the code in the "answer" box isn't the only way to solve it!  

### Exercise 1

In this exercise, you will create 3 scatter plots, one for each species of penguin, displaying penguins' body mass and flipper length. Additionally include a line of best fit.

The different parts of this exercise will walk you through the process of tidying the data and creating the scatter plot.

#### Part A

First, let's clean the data and remove any rows with missing values. Create a new dataset called `penguin_clean` which doesn't contain NA values.

```{webr-r}
#Your code here


```

::: {.callout-caution collapse="true" icon="false"}
## Check your work

To check your work, run the following code in the code chunk below.

`n_miss(penguin_clean)`

If your code works, this should print 0. 

```{webr-r}
#Copy and paste the code here 


```
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
penguin_clean<- na.omit(_____)
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
penguin_clean<- na.omit(penguin)
```
:::

#### Part B

Now use your clean data to construct a scatter plot with `flipper_length` as the explanatory variable and `body_mass` as the outcome variable.

```{webr-r}
#Your code here 


```

::: {.callout-tip collapse="true" icon="false"}
## Hint

You'll need to use ggplot, specify the data, the aesthetics (which includes the x and y variables) and then add a geometry to your plot to create points.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false

ggplot(_______, aes(x=____, y=____))+
  geom_ ______()
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
ggplot(penguin_clean, aes(x=flipper_length, y=body_mass))+
  geom_point()

#OR you could use the pipe to start with penguin_clean and then create your plot: 
#penguin_clean |> ggplot(aes(x=flipper_length, y=body_mass))+ geom_point()
```
:::

#### Part C

Now, add a line of best fit to your plot.

```{webr-r}
#Your code here 


```

::: {.callout-tip collapse="true" icon="false"}
## Hint

Add one more piece of geometry to your code from part B to your plot that creates a line of best fit.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false

ggplot(penguin_clean, aes(x=flipper_length, y=body_mass))+
  geom_point()+ 
  geom_smooth(method=_____, se=_____)

```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false

ggplot(penguin_clean, aes(x=flipper_length, y=body_mass))+
  geom_point()+ 
  geom_smooth(method="lm", se=FALSE)

#OR you could use the pipe function!
```
:::

#### Part D

Finally, let's separate the plot into 3 different plots, one for each species.

```{webr-r}
#Your code here 
 

```

::: {.callout-tip collapse="true" icon="false"}
## Hint

Use the function `facet_wrap()` and be sure to define your facet. Add this as a geometry to your code.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false

#Here's the syntax: 

facet_wrap(~_____)
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
ggplot(penguin_clean, aes(x=flipper_length, y=body_mass))+
  geom_point()+ 
  geom_smooth(method="lm", se=FALSE)+ 
  facet_wrap(~species)

#OR you could use the pipe!
```
:::

### Exercise 2

Create a bar graph which displays the number of penguins in each species. Also show which islands each penguin species is on by adjusting the fill and position of the bar graphs.

#### Part A

First, create a bar graph that shows the number of penguins in each species.

```{webr-r}
#Your code here


```

::: {.callout-tip collapse="true" icon="false"}
## Hint

`geom_bar()` is used when the data still needs to be aggregated.

`geom_col()` is used when the data is already aggregated.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
# Try this: 
penguin_clean|>
  ggplot(aes(x=_____))+
    ______
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
penguin_clean|>
  ggplot(aes(x=species))+
    geom_bar()
```
:::

#### Part B

Next, display which island each species is on by adjusting the fill of the graph.

```{webr-r}
#Your code here


#If your code works, you should see a similar bar graph to Part A, but each bar will have different colors based on the island where each species lives. 
#You should notice that two of the species live on only 1 island (not all 3). 

```

::: {.callout-tip collapse="true" icon="false"}
## Hint

This adjustment will go in the aesthetics section of your ggplot. Choose how you would like to fill the bars by using `fill = ____`.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
penguin_clean|>
  ggplot(aes(x=species, fill=______))+
  geom_bar()
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
penguin_clean|>
  ggplot(aes(x=species, fill=island))+
  geom_bar()
```
:::

#### Part C

Now that the bars are filled in by island, let's adjust them again so that they don't stack on top of each other. Instead, place all the bars next to each other.

```{webr-r}
#Your code here


```

::: {.callout-tip collapse="true" icon="false"}
## Hint

Adjust the `position` of bars in the `geom_bar()` geometry.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
penguin_clean|>
  ggplot(aes(x=species, fill=island))+
  geom_bar(position="______")
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
penguin_clean|>
  ggplot(aes(x=species, fill=island))+
  geom_bar(position="dodge")
```
:::

### Exercise 3

Create a bar graph displaying the mean bill length for each species of penguin.

#### Part A

First, we need to make our data tiny so that we can then input it into a plot. So, start by using the `summarize()` function to display the mean bill length of penguins by species.

```{webr-r}
#Your code here


#If your code works, you should see the mean bill length of each penguin species. 
```

::: {.callout-tip collapse="true" icon="false"}
## Hint

In addition to the `summarize()` function, you'll need to also use the `group_by()` function to group by species. Use the pipe to string both of these functions together.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
penguin_clean |> 
group_by(_____)|> 
  summarize(mean_bill=__________)
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
penguin_clean |> 
group_by(species)|> 
  summarize(mean_bill=mean(bill_length))
```
:::

#### Part B

Now that the data is tidy and tiny, we can plot it. Add to your code from part A and create a bar graph.

```{webr-r}
#Your code here


```

::: {.callout-tip collapse="true" icon="false"}
## Hint

Recall: `geom_bar()` is used when the data still needs to be aggregated.

`geom_col()` is used when the data is already aggregated.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
penguin_clean |> 
  group_by(species)|> 
  summarize(mean_bill=mean(bill_length)) |>
  ggplot(aes(x=____, y=_____)) +
  geom_col()
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
penguin_clean |> 
  group_by(species)|> 
  summarize(mean_bill=mean(bill_length)) |> 
  ggplot(aes(x=species, y=mean_bill)) +
  geom_col()
```
:::

#### Part C

Finally, we want to order the columns in the bar graph from smallest to largest. To do this, we'll use `fct_reorder()`. Find where it goes in your current code. Hint: You will not have to write a new line.

Feel free to also add other features to your graph, like a title, colored bars, etc!

```{webr-r}
#Your code here


```

::: {.callout-tip collapse="true" icon="false"}
## Hint

What are you trying to reorder on the graph? Are you trying to reorder the x variable or y variable?
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
penguin_clean |> 
  group_by(species)|> 
  summarize(mean_bill=mean(bill_length)) |> 
  ggplot(aes(x=fct_reorder(_____, _____), y=mean_bill))+
  geom_col()
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
penguin_clean |> 
  group_by(species)|> 
  summarize(mean_mass=mean(body_mass))|>
  ggplot(aes(x=fct_reorder(species, mean_mass), y=mean_mass))+
  geom_col()
```
:::

### Exercise 4

Create a density plot of penguin flipper length. Include a line displaying the mean flipper length. Additionally, separate the data by island, so that you have one density curve per island.

#### Part A

First, create a density curve of penguin flipper length.

```{webr-r}
# Your code here 


```

::: {.callout-tip collapse="true" icon="false"}
## Hint

Use the geometry command `geom_density()`.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
ggplot(_______, aes(x=______))+
  _________

#OR you can use the pipe: 

_____ |> 
  ggplot(aes(x=_____))+ 
    ________
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
ggplot(penguin_clean, aes(x=flipper_length))+
  geom_density()

#OR you can use the pipe: 

#penguin_clean|>
#ggplot( aes(x=flipper_length))+
#  geom_density()

```
:::

#### Part B

Next, separate out your density curve and create one density curve for each species. Hint: You will not need to add a whole new line of code to do this, just add a command into your existing code.

```{webr-r}
# Your code here


```

::: {.callout-tip collapse="true" icon="false"}
## Hint

To separate the plot by island, you will need to adjust the color in the aesthetics section of `ggplot`.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false

ggplot(penguin_clean, aes(x=flipper_length, color=______))+
  geom_density()

```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false

ggplot(penguin_clean, aes(x=flipper_length, color=island))+
  geom_density()

```
:::

#### Part C

Next, we'll add a line displaying the mean flipper length for each island. To do this, we need to first create tidy, tiny data which displays the mean flipper lengths.

Create a new variable called `summary_data` which displays a chart with the mean flipper length for each island.

```{webr-r}
#Your code here: 


 
```

::: {.callout-caution collapse="true" icon="false"}
## Check your work

To check your work, run the following code in the code chunk below.

`summary_data`

If your code is successful, this will print the mean flipper length for each island.

```{webr-r}
#Copy and paste the code here 


```
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

You'll need to use the following functions in your code: `group_by()`, `summarize()`, and `mean()`.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
summary_data <- ________ |> 
  group_by(_____)|> 
  summarize(_________)
```
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
summary_data <- penguin_clean |> 
  group_by(______)|> 
  summarize(mean_flipper = mean(______))
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
summary_data<- penguin_clean|> 
  group_by(island)|>
  summarize(mean_flipper=mean(flipper_length))
```
:::

#### Part C

Finally let's add vertical lines to the graph indicating the mean flipper length for each island.

To do this, use the geometry `geom_vline()`. Use your `summary_data()` to specify the conditions in `geom_vline()`.

```{webr-r}
#Your code here 


#Your final plot should include 3 density lines and 3 vertical lines indicating the mean for each island. 
```

::: {.callout-tip collapse="true" icon="false"}
## Hint

First define your `summary_data` and `mean_flipper`, like you did in Part B.

Then create the ggplot, like you did in Part A. Add a `geom_vline()` to the ggplot.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

The syntax for `geom_vline()` is `geom_vline(data=____, aes(xintercept=_____, color=_____))`.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
summary_data<- penguin_clean|> 
  group_by(island)|>
  summarize(mean_flipper=mean(flipper_length))

ggplot(penguin_clean, aes(x=flipper_length, color=island))+
  geom_density()+
  geom_vline(data=_______, aes(xintercept=_______, color=______))
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
summary_data<- penguin_clean|> 
  group_by(island)|>
  summarize(mean_flipper=mean(flipper_length))

ggplot(penguin_clean, aes(x=flipper_length, color=island))+
  geom_density()+
  geom_vline(data=summary_data, aes(xintercept=mean_flipper, color=island))

```
:::

Congratulations, you have finished this tutorial!!
