---
format: html
editor: visual
filters: 
  - webr
title: "Missing Data Tutorial"
---

```{webr-r}
#| context: setup
#| label: input data
#| tidy:true

library(tidyverse)
library("palmerpenguins")
library("naniar")

penguin_eye<- c('X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','brown','brown','red','grey','grey','grey','brown','brown','red','grey','brown','brown','red','grey','brown','brown','red','grey','red','red','brown','brown','red','grey','brown','brown','red','brown','brown','red','grey','brown','brown','brown','red','grey','grey','grey','brown','brown','red','grey','brown','brown','red','grey','brown','brown','red','grey','red','red','brown','brown','red','grey','brown','brown','red','brown','brown','red','grey','brown','brown','brown','red','grey','grey','grey','brown','brown','red','grey','brown','brown','red','grey','brown','brown','red','grey','red','red','brown','brown','red','grey','brown','brown','red','brown','brown','red','grey','brown','brown','brown','red','grey','grey','grey','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','red','grey','brown','brown','red','grey','red','red','brown','brown','red','grey','brown','brown','red','brown','X','red','grey','brown','brown','red','grey','red','red','brown','brown','red','grey','brown','brown','red','brown','X','red','grey','red','red','brown','brown','red','grey','brown','brown','red','brown','red','grey','red','red','brown','brown','red','grey','brown','brown','red','brown','red','grey','red','red','brown','brown','red','grey','brown','brown','red','brown','grey','red','red','brown','X','red','grey','brown','brown','red','brown','red','grey','red','red','brown','brown','red','grey','brown','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','X','red','red','brown','brown','red','grey','brown','brown','red','brown','red','grey','red','red','brown','brown','red','red','brown','brown','red','red','brown','brown','brown','brown','red','grey','brown','brown','brown','brown','red','grey','brown','brown','red','brown','red','grey','brown','X')

penguin<- tibble('species'=penguins$species, 'island'=penguins$island, 'body_mass'=as.numeric(penguins$body_mass_g), 'year'=penguins$year, 'sex'=penguins$sex, 'eye_color'=penguin_eye)

#knitr::opts_chunk$set(error = TRUE)
#knitr::opts_chunk$set(echo = TRUE)
```

## Welcome

In this tutorial, you will learn how to deal with missing data using the `naniar` package. The `naniar` package helps visualize where there are missing or 'NA' values in a dataset. It allows you to see how many 'NA' values there are and where they show up.

### Functions in Focus

Let's start by looking at some of our key functions which we'll use in this exercise.

1.  `na_if()`: Converts values to `NA`. Can be combined with the `mutate()` function to create or replace a column with `NA` values.

Syntax: `na_if(vector, value to replace)`

2.  `replace_na()`: Changes an `NA` value to a given value.

Syntax: `replace_na(vector, value)`

3.  `miss_var_summary()`: Creates a summary table of missing values by variable. Includes the *number* of missing values per variable (called `n_miss`) and the *percentage* of missing values per variable (called `pct_miss`).

Syntax when used with a dataframe: `miss_var_summary(df)`

4.  `gg_miss_var()`: Creates a bar graph of missing values by variable.

Syntax: `gg_miss_var(df)`

-   You can also choose one variable and create a different graph for each aspect of that variable using the syntax `gg_miss_var(df, facet=variable)` where `variable` is your chosen variable.

5.  `vis_miss()`: Creates a visual of missing values by variable by position in the dataset.

Syntax: `vis_miss(df)`

## Our Dataset

We will work with a dataset with measurements for 3 different species of penguins on islands in the Palmer Archipelago, Antarctica. We'll be focusing on the measurements for species, location, body mass (g), sex, and year the data was taken. For this tutorial, we've also added a new column called `eye_color`. Here's more [information](https://cran.r-project.org/web/packages/palmerpenguins/readme/README.html) about the [data set](https://cran.r-project.org/web/packages/palmerpenguins/readme/README.html)!

Here is a preview of the data you will be working with:

```{webr-r}
penguin
```

Notice that there are NA values as well as "X"s in our data set. Today we'll be looking at those NA (or missing) values.

## Exercises

Here are some practice exercises. You can click for hints and the answer for each problem.

### Exercise 1

First, let's deal with the "X" values in our data set. We assume that these are missing values, so we want to replace the "X" with an "NA" so that we can use the `naniar` package functions on those values.

Use the `na_if()` function to replace all the X's with `NA` values.

Rewrite your data frame `penguin` so that when we pull up `penguin` in the future, all the "X" values will be gone.

```{webr-r}
#Your code here



```

::: {.callout-caution collapse="true" icon="false"}
## Check your work

To check your work, run the following code in the code chunk below.

`penguin`

If your code works, this should display your new data set with NA instead of X. 

```{webr-r}
#Copy and paste the code here 


```
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

Use the pipe function to specify the `penguin` data set and then use the `mutate()` function to mutate the column `eye_color`.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
penguin |> mutate(____=na_if(_____, '____'))
```
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

Once you've mutated the `eye_color` column, you need to rewrite the `penguin` data frame with that new column. Use the variable assignment function (`<-`) to do this.
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
penguin <- penguin |> mutate(eye_color=na_if(eye_color, 'X'))
```
:::

### Exercise 2

Now, let's find where we have `NA` values in our code. Use the command `miss_var_summary()` to find which variables have missing values and how many missing values each variable contains.

```{webr-r}
#Your code here


#If your code works, you should see there are 114 missing observations in the eye color variable, which is 33% of the all the eye color observations.
#You should also see that 3 variables have no missing data. 
```

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
miss_var_summary(________)

```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
miss_var_summary(penguin)

```
:::

### Exercise 3

You may have noticed that `eye_color` has the highest number of missing values, so we'll focus on the missing values in `eye_color` for this exercise.

Use `miss_var_summary()` to display the number of missing `eye_color` values, separated into individual years. Don't include any other variables.

```{webr-r}
#Your code here 


#If your code works, you should see the number of missing eye_color values for each year.
#You'll see that one year contains only missing values. 

```

::: {.callout-tip collapse="true" icon="false"}
## Hint

Use the pipe function to select the specific data you want to display.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

Use the `select()` function to choose the variables you want to focus on and use the `group_by()` function to choose how you want to layout the chart.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
penguin |> 
  select(___, ____) |>
  group_by(_____) |>
  miss_var_summary()

```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
penguin |> 
  select(year, eye_color) |>
  group_by(year) |>
  miss_var_summary()

```
:::

### Exercise 4

Next use `gg_miss_var()` to create bar graphs that show the number of missing values per variable, by island. Create 3 graphs total, 1 graph for each island.

```{webr-r}
# Your code here: 

#If your code worked, you should see 3 different missing value graphs titled "Biscoe" "Dream" and "Torgersen". 
```

::: {.callout-tip collapse="true" icon="false"}
## Hint

Recall the syntax `gg_miss_var(df, facet=variable)` where `variable` is your chosen variable.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
gg_miss_var(penguin, facet=____)
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
gg_miss_var(penguin, facet=island)
```
:::

### Exercise 5

Next, let's look at where the missing values are showing up within the observations in the dataset.

First, arrange the data by year, then input it into the `vis_miss()` function to see where the `NA` values fall when the data is sorted by year.

```{webr-r}
# Your code here


# If your code works, it will create a chart with lines marking the locations of `NA` vales.

#You should see black lines for all the NA values, 
#and a gray background representing observations which do NOT contain NA values. 

#The chart should be sorted by year, so there should be a black box in the eye_color column that corresponds to the year 2007 when all the values are NA. 
```

::: {.callout-tip collapse="true" icon="false"}
## Hint

Use the pipe function (`|>`) with the `arrange()` function to sort the data by year.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
penguin |>
  arrange(____) |>
  _______(_____)
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
penguin |>
  arrange(year) |>
  vis_miss()
```
:::

### Exercise 6

Finally, let's imagine that you don't want any `NA` values in the `body_mass` column. Use `replace_na()` to convert all of the `NA` values in that column to the mean of the `body_mass` column.

```{webr-r}
#Your code here


#If your code works, you should not see any `NA` values in the body_mass column. Instead of an NA in row 4, there should be a 4202. 
```

::: {.callout-tip collapse="true" icon="false"}
## Hint

Use the `mean()` function to find the mean of `body_mass`.

Use the pipe to specify the penguin data and the `mutate()` function to change the `body_mass()` column and the `replace_na()` function to convert the NA values to the mean.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
penguin |>
  mutate(____ = replace_na(_____, mean(_____, na.rm=TRUE)))
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
penguin |> mutate(body_mass=replace_na(body_mass, mean(body_mass, na.rm=TRUE)))
```
:::

Congratulations, you finished this tutorial!
