---
format: html
editor: visual
filters: 
  - webr
title: "Data Transformation in R"
---

```{webr-r}
#| context: setup
#| label: input data and seeds

library(tidyverse)

set.seed(123)  # Setting seed for reproducibility
climate_data_wider <- tibble(
  country = rep(c("Freedonia", "Sylvania", "Zubrowka", "Borduria"), each = 4),
  year = rep(2020:2023, 4),
  carbon_emission = round(runif(16, 4, 12), 2), # In metric tons per capita
  temperature_change = round(runif(16, 0.5, 3.0), 2), # In degree Celsius compared to the 20th-century average
  treaty_participation = sample(c(TRUE, FALSE), 16, replace = TRUE)
)

climate_data_longer <- climate_data_wider |>
  pivot_longer(cols = c(carbon_emission, temperature_change), 
               names_to = "measurement", 
               values_to = "value")

# Additional data for join examples
treaty_data <- tibble(
  country = c("Freedonia", "Sylvania", "Zubrowka", "Elbonia"),
  treaty_signed = c(TRUE, TRUE, FALSE, TRUE)
)

new_data <- tibble(
  country = rep("Elbonia", 4),
  year = 2020:2023,
  carbon_emission = round(runif(4, 4, 12), 2),
  temperature_change = round(runif(4, 0.5, 3.0), 2),
  treaty_participation = sample(c(TRUE, FALSE), 4, replace = TRUE)
)



knitr::opts_chunk$set(error = TRUE)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

In this tutorial, we'll explore several critical tidverse functions for data transformation with environmental data. The data represents fictional measurements from different countries concerning carbon emissions, temperature changes, and participation in various environmental treaties.

Here are the functions you will need to use:

1.  `bind_rows()`: This function is used to combine two or more data frames by row, binding them together. Syntax: `combined_data <- bind_rows(df1, df2)`

2.  `pivot_longer`: Reshape data when variables are spread across columns. Syntax: `pivot_longer(data, cols_to_pivot, names_to="name", values_to="value")`

3.  `pivot_wider`: Reshape data when you need to spread rows across new columns. Syntax: `pivot_wider(data, names_from = "category", values_from = "amount")`

4.  `left_join()`: Joins two datasets, retaining observations from the first (left) dataframe. Syntax: `left_join(df1, df2, by = "key_variable")`

5.  `anti_join()`: Identifies rows in one data frame that do not have a corresponding match in another data frame. Syntax: `result <- anti_join(df1, df2, by = "key_variable")`

## Our Dataset

Let's create our synthetic dataset first.

This dataset includes fictional information about carbon emissions and temperature changes over four years for various countries, as well as whether they participated in environmental treaties. Here is your first dataeframe, called `climate_data_longer`:

```{webr-r}

climate_data_longer

```

## Exercises

Here are some practice exercises. You can check if your code works using the "check your work" button after each problem. 

If you get stuck, you can also click for hints and for an example answer. Remember, there's many ways to solve each problem, so the code in the "answer" box isn't the only way to solve it!  

### Exercise 1: Widen the Data

This data is too long, with multiple variables sharing the same columns. Convert `climate_data_longer` to a wider format, where each column represents a different type of measurement. Use the current measurement variable to create new variable names. Name the wider dataframe `climate_data`.

```{webr-r}
# Your code here.  


```

::: {.callout-caution collapse="true" icon="false"}
## Check your work

To check your work, run the following code in the code chunk below.

`str(climate_data)`

If run successfully, the following code would indicate that your data has 16 observations and 5 variables

```{webr-r}
#Copy and paste the code here 


```
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

You'll want to use the `pivot_wider()` function. You will need to specify the names_from and values_from parameters.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
# Here's how you start:
climate_data <- climate_data_longer |> 
  pivot_wider(names_from = _______, values_from = ______)
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
climate_data <- climate_data_longer |> 
  pivot_wider(names_from = measurement, values_from = value)
```
:::

### Exercise 2: Lengthen the Data

For practice purposes, can you convert `climate_data` back to its original, longer format, where one column represents the type of measurement (carbon emission or temperature change), and another column represents the values? Call this new data frame `climate_data_reversion`.

```{webr-r}
# Your code here

 
```

::: {.callout-caution collapse="true" icon="false"}
## Check your work

To check your work, run the following code in the code chunk below.

`str(climate_data_reversion)`

If run successfully, the following code should show 32 observations, 5 variables.

```{webr-r}
#Copy and paste the code here 


```
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

This time, you'll want to use the pivot_longer() function. You need to specify which columns you want to make longer.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false 
# Start with this:
climate_data_reversion <- climate_data |> 
  pivot_longer(cols = c(______, _____), 
              names_to = "______", 
              values_to = "______")
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
climate_data_reversion <- climate_data |> 
  pivot_longer(cols = c(carbon_emission, temperature_change),
               names_to = "measurement", 
               values_to = "value")
```
:::

### Exercise 3: Adding Rows

Suppose we have measured data for another country, "Elbonia," and want to add this data to our existing `climate_data`. Use `bind_rows()` to do this.

First, we will create the new data with the following code:

```{webr-r}
new_data <- tibble(
  country = rep("Elbonia", 4),
  year = 2020:2023,
  carbon_emission = round(runif(4, 4, 12), 2),
  temperature_change = round(runif(4, 0.5, 3.0), 2),
  treaty_participation = sample(c(TRUE, FALSE), 4, replace = TRUE)
)

new_data
```

Now, bind this new data to `climate_data` and call your new data set `all_climate_data`

```{webr-r}
# Your code here


```
::: {.callout-caution collapse="true" icon="false"}
## Check your work

To check your work, run the following code in the code chunk below.

`str(all_climate_data)`

If run successfully, the following code should now show 20 observations.

```{webr-r}
#Copy and paste the code here 


```
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

Use bind_rows() to combine the rows of two datasets. Make sure they have the same columns!
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
# Here's how you start:
all_climate_data <- bind_rows(____, _____)
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
all_climate_data <- bind_rows(climate_data, new_data)
```
:::

### Exercise 4: Joining Data

We also have another dataset which tells us which countries have signed an environmental treaty. The dataset is called `treaty_data`.

Let's find out which countries in our `climate_data` have signed an environmental treaty using `left_join()`.

```{webr-r}
# Your code here

 
```

::: {.callout-caution collapse="true" icon="false"}
## Check your work

To check your work, run the following code in the code chunk below.

`str(climate_treaty_info)`

If run successfully you should have more than 5 variables.

```{webr-r}
#Copy and paste the code here 


```
:::
::: {.callout-tip collapse="true" icon="false"}
## Hint

You can use `left_join()` to merge data, keeping all records from `climate_data` and joining information from `treaty_data`.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
# Here's how you start:
climate_treaty_info <- left_join(______, ______, by = "______")
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
climate_treaty_info <- left_join(climate_data, treaty_data, by = "country")
```
:::

### Exercise 5: Anti-Join

Now, identify which countries listed in `treaty_data` are not present in `climate_data` using `anti_join()`. Save the mismatched countries in an object called `missing_countries`.

```{webr-r}
# Your code here

 
```
::: {.callout-caution collapse="true" icon="false"}
## Check your work

To check your work, run the following code in the code chunk below.

`str(missing_countries)`

If run successfully this should have 1 observation, Elbonia.

```{webr-r}
#Copy and paste the code here 


```
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

anti_join() will return rows from x where there are no corresponding rows in y that match by the given key.
:::

::: {.callout-tip collapse="true" icon="false"}
## Hint

```{r}
#| eval: false
# Here's how you start:
missing_countries <- anti_join(______, ______, by = "_______")
```
:::

::: {.callout-note collapse="true" icon="false"}
## Answer

```{r}
#| eval: false
missing_countries <- anti_join(treaty_data, climate_data, by = "country")
```
:::

## Conclusion

Great work! In this tutorial, you've learned how to use various tidyverse functions to manipulate a climate change dataset. These functions are versatile and can be applied to many data manipulation tasks in R. Keep practicing and exploring their different parameters and capabilities.
